Describing demographic models is hard

### 2020-05-18 paper.tex
Some comments from Dominic on an early draft

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}

@@

The \msprime\ coalescent simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
is now quite widely used. The large increases in efficiency over the classical
\ms\ program~\citep{hudson2002generating} make it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces. In particular, programs like \ms\
required users to specify cryptic command line options to describe demographic
models [JK: maybe include an example ms command line?]. Particularly for
the large models we are using today, these are not comprehensible for humans.
The Python API for \msprime\ is a great improvement, allowing the user to
state models in a documented and reproducible manner. [JK: maybe show
the same model as described above in msprime notation?]

Implementing multi-population models of demographic history is still hard and error
prone. In this note we discuss how two poor design decisions
in \msprime's demography API lead to errors being made, which ended
up in the scientific record. 
\aprcomment{You could argue it is poor API design, and I see that.. but I also
think there is something more to be said about best practices in implementing simulations, 
including your own various verifications and sanity checks, and the need for robust
implementation of commonly used models (stdpopsim). I do think that in the end
the responsibility lies with the user in making sure that everything is implemented
properly.}

@@

This model was subsequently copied
several times and used in papers [JK: do we want to estimate how
many times it was copied? A quick search on github suggests around 30
times].

Arguably, this error occurred because of a poor API design choice.
Currently, to model a population split, the user must use a MassMigration event to move 
lineage from one population to another and then also remember to turn off migration
between those populations at the same time.
The release of \msprime\ 1.0 will introduce a \texttt{PopulationSplit} event, which allows
such models to be described more declaratively.

@@

These errors suggests three lessons. 
First, the msprime API design was poorly designed, requiring the user to pass 
three separate parameters to specify a demographic model. 
In \msprime\ 1.0 we introduce
a Demography class, which wraps these three parameters.
Thus, instead of writing
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
\end{lstlisting}
we would now write
\begin{lstlisting}[frame=single]
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}

@@

Defining demographic models is error prone, and incorrect implementation can  
impact downstream analysis and interpretation. The errors that we highlighted here have 
exposed some weaknesses of the current \msprime\ API, which will be fixed in an 
upcoming release. Recognition of these difficulties is also a 
primary motivating factor in efforts to standardize and vet demographic model 
implementation in \stdpopsim~\cite{adrion2019community}, with quality-controlled 
models and simulation resources for a growing number of commonly studied species.

If you do need to implement your own demographic model, we recommend that steps are taken to 
verify its correctness. A second pair of eyes (code review) or independent implementation 
can go a long way in catching subtle bugs (this is the approach taken by \stdpopsim). It is also 
very important to perform some basic statistical verification of the simulation, such as checking 
output heterozygosity and $F_{ST}$ against expectations. If simulations are large so that 
statistical validation is burdensome, you can instead simulate over a smaller genome length 
and with fewer individuals to make sure basic properties hold. For example, a routine statistical 
checks of $F_{ST}$ would surely have caught the error in~\citet{martin2017human}. 
\aprcomment{too snarky...? but I think it's an important point}

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time.
    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct 
    demographic model and found that risk prediction was greatly improved over what was originally 
    reported in all three populations.
    \aprcomment{directly compare to figure 5 in the original paper}
    \aprcomment{to do: fill in caption}
}
\label{fig:prs}
\end{center}
\end{figure}

>>>>

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\dncomment}[1]{{\textcolor{red}{Dom: #1}}}

@@

The \msprime\ coalescent simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
is now quite widely used. The large increases in efficiency over the classical
\ms\ program~\citep{hudson2002generating} make it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces. In particular, programs like \ms\
required users to specify cryptic command line options to describe demographic
models [JK: maybe include an example ms command line?]. Particularly for
the large models we are using today, these are not comprehensible for humans.
\dncomment{That's a pretty strong statement, might be good to give an example 
or describe what it would look like (``specifying the gravel ooa model would take 
10K lines of parameter specification''), or just say that its much less intuitive and 
so more error prone}
The Python API for \msprime\ is a great improvement, allowing the user to
state models in a documented and reproducible manner. [JK: maybe show
the same model as described above in msprime notation?]

Implementing multi-population models of demographic history is still hard and error
prone. In this note we discuss how two poor
\dncomment{Overly harsh I think, could drop this word} design decisions
in \msprime's demography API lead
\dncomment{Also not fair I don't think. I mean Alicia's bug hints at a lack of validation that 
could have happened at any stage in the pipeline, and did happen elsewhere 
I believe. Could make this more positive by describing how these design choices 
could be improved to reduce the chance of errors going unnoticed/eliminate 
some entirely} to errors being made, which ended
up in the scientific record. 
\aprcomment{You could argue it is poor API design, and I see that.. but I also
think there is something more to be said about best practices in implementing simulations, 
including your own various verifications and sanity checks, and the need for robust
implementation of commonly used models (stdpopsim). I do think that in the end
the responsibility lies with the user in making sure that everything is implemented
properly.}

@@

This model was subsequently copied
several times and used in papers [JK: do we want to estimate how
many times it was copied? A quick search on github suggests around 30
times]. \dncomment{I like this as a rough metric, like ``the number of copies 
is unknown, but we found x copies on github''}

Arguably, this error occurred because of a poor 
\dncomment{Again I'd go positive. If msprime is so bad, god help the rest of us. 
Something like ``To model a population split in the original API, ...''?}
API design choice.
Currently, to model a population split, the user must use a MassMigration event to move 
lineage from one population to another and then also remember to turn off migration
between those populations at the same time.
The release of \msprime\ 1.0 will introduce a \texttt{PopulationSplit} event, which allows
such models to be described more declaratively. 
\dncomment{Strikes me as a bit programmer-centric. Maybe intuitively? Simply?}

@@

These errors suggests three lessons. 
First, the msprime API design was poorly designed, 
\dncomment{Still don't like this word. Again if msprime is poor then most bioinformatics 
software is a dumpster fire. What about ``...msprime API can be adapted to 
simplify common use-cases''? A bit clunky but maybe along those lines
}
requiring the user to pass 
three separate parameters to specify a demographic model. 
In \msprime\ 1.0 we introduce
a Demography class, which wraps these three parameters.
Thus, instead of writing
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
\end{lstlisting}
we would now write
\begin{lstlisting}[frame=single]
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}

@@

Defining demographic models is error prone \dncomment{Just a thought, but alternatively could say 
``Demographic models are complex''}, and incorrect implementation can  
impact downstream analysis and interpretation. The errors that we highlighted here have 
exposed some weaknesses of the current \msprime\ API, which will be fixed in an 
upcoming release. Recognition of these difficulties is also a 
primary motivating factor in efforts to standardize and vet demographic model 
implementation in \stdpopsim~\cite{adrion2019community}, with quality-controlled 
models and simulation resources for a growing number of commonly studied species.

If you do need to implement your own demographic model, we recommend that steps are taken to 
verify its correctness. A second pair of eyes (code review) or independent implementation 
can go a long way in catching subtle bugs (this is the approach taken by \stdpopsim). It is also 
very important to perform some basic statistical verification of the simulation, such as checking 
output heterozygosity and $F_{ST}$ against expectations. If simulations are large so that 
statistical validation is burdensome, you can instead simulate over a smaller genome length 
and with fewer individuals to make sure basic properties hold. For example, a routine statistical 
checks of $F_{ST}$ would surely 
\dncomment{I'd go with either this or 'routine' but maybe not both
}
have caught the error in~\citet{martin2017human}. 
\aprcomment{too snarky...? but I think it's an important point}

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time.
    \dncomment{Might have been on your todos below but would help I think to have the correct model 
    drawn out too (see now that you did in fig 1. I'd at least point to that to highlight the difference)}
    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct 
    demographic model and found that risk prediction was greatly improved over what was originally 
    reported in all three populations.
    \aprcomment{directly compare to figure 5 in the original paper}
    \aprcomment{to do: fill in caption}
}
\label{fig:prs}
\end{center}
\end{figure}
